---
title: "Coronavirus Application R Package"
description: |
  This R package contants the application that explores the current state of coronavirus made for ETC5523: Communicating with Data
author:
  - name: Vinny Vu 
    url: https://www.linkedin.com/in/vinny-vu-809bb1139/
date: 10-23-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating blogs with Distill at:
# https://rstudio.github.io/distill/blog.html

```

This package contains the shiny application and functions exploring the current state of coronavirus as at 31 July 2020.


Follow the link to go to the pkgdown website for more information: [CoronaShiny](https://etc5523-2020.github.io/r-package-assessment-vvuu0009/)


## Installation

<!-- You can install the released version of CoronaShiny from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("CoronaShiny") -->
<!-- ``` -->

The development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("etc5523-2020/r-package-assessment-vvuu0009")
```


The main functions in this package includes:

- **launch_app**: This function runs the Coronavirus Shiny Application exploring the state of Coronavirus as at July 31 2020  
- **cumplot**: This function produces a line graph of coronavirus cases for a selected case type and continent with date on the x-axis and daily cases for the selected `Type` and `Continent` on the y-axis.
- **selectInput01**: This function changes the selectInput function of the Shiny app based on the desired input 

This package also contains the following data sets:

- **corona_cont**: This data set shows the daily coronavirus cases by each continent by case type
- **coronavirus_july20**: This data set shows the cumulative coronavirus cases as at 31 July 2020 for each country 

## Example

Below displays a sample of both data sets:

```{r corona_cont}
library(CoronaShiny)
library(tibble)
corona_cont
```

```{r coronavirus_july20}
coronavirus_july20
```

Below displays an example cumplot for confirmed cases in Asia

```{r}
cumplot("confirmed", "Asia")
```